---
title: "Reproducible Research PA1"
author: "Arunkumar"
date: "28/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Introduction

The data for this assignment can be downloaded from the link below:

Dataset: Activity monitoring data 
Download Link: http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

#Dataset Variables in the file:

- steps: Number of steps taking in a 5-minute interval (NA used to show missing values)
- date: The date on which the measurement was taken in YYYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken
- dataset is stored as a comma-separated-value (CSV) file
- total of 17,568 observations in this dataset

#Code for reading in the dataset and/or processing the data

```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
  temp <- tempfile()
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
  unzip(temp)
  unlink(temp)
}
data <- read.csv("activity.csv")
```


# Histogram of the total number of steps taken each day



```{r}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Per Day"), col="orange", xlab="Number of Steps")
png("histogram_totalsteps.png")
```
# Mean and median number of steps taken each day


```{r}
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

The Mean = 10766.19 and The median = 10765


# Time series plot of the average number of steps taken
- Calculate average steps for each interval for all days.
- Plot the Average Number Steps per Day by Interval.
- Find interval with most average steps.


```{r}
steps_by_interval <- aggregate(steps ~ interval, data, mean)
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="No. of Steps",main="Avg. No. of Steps per Day by Interval" )
```

# The 5-minute interval that, on average, contains the maximum number of steps

```{r}
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

5-minute interval, contains the maximum number of steps = 835


#Code to describe and show a strategy for imputing missing data

Missing values were imputed with by inserting the average for each interval. for instance, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA.

```{r}
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
```

The first day (10-01-2012) was imputed with 0 as it would will be over 9,000 steps higher than the following day, which had only 126 steps. 

```{r}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```
# Histogram of the total number of steps taken each day after missing values are imputed

Recount total steps by day and create Histogram.

```{r}
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps")
 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="orange", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("red", "orange"), lwd=10)
```

Create Histogram to show difference.



Calculate new mean and median for imputed data.

```{r}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

Calculate difference between imputed and non-imputed data.

```{r}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

Calculate total difference.

```{r}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```

*The imputed data mean is 1.059 × 104
*The imputed data median is 1.0766 × 104
*The difference between the non-imputed mean and imputed mean is -176.4949
*The difference between the non-imputed mean and imputed mean is 1.1887
*The difference between total number of steps between imputed and non-imputed data is 7.5363 × 104. Thus, there were 7.5363 × 104 more steps in the imputed data.
#**Are there differences in activity patterns between weekdays and weekends?**

Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.

```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday" )
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))
steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)
library(lattice)
xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Avg. Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l") 
```
